\documentclass[11pt]{article}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[authoryear,round]{natbib}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

\title{\textbf{Product Geometry for Predictable Linear Reinforcement}\\
Elephant Walks, Universality and Critical Windows}

\author{Research draft}
\date{\today}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

\newcommand{\F}{\mathcal{F}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Pbb}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Law}{\mathcal{L}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\D}{\mathcal{D}}

\begin{document}
\maketitle

\begin{abstract}
We study discrete-time processes whose conditional drift has the predictable
linear reinforcement form
\[
  \E[\xi_{n+1}\mid\F_n]=a_nS_n/n,\qquad S_n=\sum_{k\le n}\xi_k,
\]
with bounded predictable environment \(a_n\). The point is not a new variant of
one random walk, but a classification principle. Once the environment is fixed,
the asymptotic behaviour is governed by two canonical quantities: the
multiplicative normalisation
\[
  A_n=\prod_{k<n}\left(1+\frac{a_k}{k}\right)
\]
and the quadratic variation of the martingale \(S_n/A_n\). The main theorem
proves a product--bracket classification for predictable linear reinforcement
schemes satisfying explicit regularity and Lindeberg conditions.
Infinite bracket gives Gaussian martingale limits with deterministic time
changes determined by the ratios of the product and bracket; finite bracket
gives a terminal martingale law. Predictable elephant random walks are then a
canonical application: their bracket decomposes as a product-determined clock
minus a negligible self-interaction term. This yields the classical diffusive,
critical and superdiffusive regimes as corollaries and exposes logarithmic
critical windows in which \(a_n\to1/2\) produces three different normalising
scales. We also prove rigidity: schemes with equivalent product and bracket
geometry have the same scaling limits. Counterexamples show why convergence of
the coefficient alone cannot classify the asymptotics.
\end{abstract}

\noindent\textbf{Keywords:} predictable reinforcement; elephant random walk;
martingales; universality; phase transition; regular variation.

\section{Introduction}

Many reinforced and self-interacting processes have a linear first-order drift:
after a suitable projection, the next increment has conditional mean
proportional to the present state divided by time. In the simplest scalar form,
\[
  \E[\xi_{n+1}\mid\F_n]=a_n\frac{S_n}{n},
  \qquad S_n=\sum_{k\le n}\xi_k,
\]
where the coefficient \(a_n\) is predictable. Fixed \(a_n\) gives the classical
linear reinforcement exponents. Predictable \(a_n\), however, creates a
different problem: small changes in the drift accumulate over logarithmic time,
so the limiting coefficient alone cannot determine the asymptotics.

This paper isolates the invariant behind this phenomenon. The conditional drift
is removed by the multiplicative product
\[
  A_n=\prod_{k<n}\left(1+\frac{a_k}{k}\right),
\]
and the remaining fluctuations are governed by the quadratic variation of the
martingale \(S_n/A_n\). The product and the bracket are therefore the two
canonical coordinates of a predictable linear reinforcement scheme. The main
result proves that, under explicit regularity and Lindeberg conditions, these
coordinates classify the scaling limits: infinite bracket gives Gaussian limits
with deterministic time change, while finite bracket gives a terminal martingale
law.

The elephant random walk enters as the basic complete-memory model where this
principle can be pushed to a concrete phase diagram. At time \(n+1\), the walker
chooses a past increment uniformly and either repeats or reverses it. For a fixed
memory parameter the model has diffusive, critical and superdiffusive regimes
\citep{schutz2004elephants,colette2017central,baur2016elephant,bercu2018martingale}.
If the repeat--reverse coefficient is predictable, the same martingale product
appears, but the bracket contains a self-interaction correction. Showing that
this correction is negligible in the appropriate regimes turns the abstract
classification into explicit ERW limits.

The point is not that \(p\) has been replaced by \(p_n\). The point is that the
phase boundary is product-geometric. For example,
\[
  a_n=\frac12+\frac{c}{\log n}
\]
has the same pointwise limit \(1/2\) for every \(c\), but it produces three
different bracket geometries and hence three different normalising scales. This
logarithmic critical window is the simplest place where the hidden geometry is
visible: convergence of the environment is too crude, while the product and
bracket see the phase transition.

This perspective separates the present work from several neighbouring theories.
Classical ERW methods rely on a fixed exponent. Urn methods rely on a fixed or
asymptotically stable replacement scheme
\citep{janson2004urns,janson2005triangular,laruelle2013urns}. Reinforced
processes exhibit many memory mechanisms \citep{pemantle2007survey,laulin2022reinforced},
and stochastic approximation studies adaptive parameters through their own
limiting recursion \citep{kushner2003stochastic}. Here the object being
classified is the predictable reinforcement environment itself, through the
product--bracket geometry it induces.

The paper is organised as follows. Section~\ref{sec:model} defines predictable
linear reinforcement schemes and then specialises them to predictable elephant
walks. Section~\ref{sec:classification} states the master classification theorem
and the rigidity principle. Section~\ref{sec:proofs} proves the martingale
representation and the classification. Section~\ref{sec:regimes} derives the
ERW phase diagram. Section~\ref{sec:optimality} gives counterexamples and the
logarithmic critical window. Section~\ref{sec:examples} records learning rules
only as examples of predictable environments, and Section~\ref{sec:open}
collects open problems.

\section{Predictable Linear Reinforcement}\label{sec:model}

Let \((\Omega,\F,\Pbb)\) carry a filtration \((\F_n)_{n\ge0}\). A scalar
predictable linear reinforcement scheme is an adapted process
\[
  S_n=S_0+\sum_{k=1}^n \xi_k
\]
with square-integrable increments such that, for a bounded predictable
environment \((a_n)_{n\ge1}\),
\[
  \E[\xi_{n+1}\mid\F_n]=a_n\frac{S_n}{n},
  \qquad n\ge1.
\]
Throughout we assume that, for some \(\varepsilon\in(0,1)\),
\[
  -1+\varepsilon\le a_n\le 1-\varepsilon.
\]
This is the natural non-degeneracy condition for the multiplicative
normalisation. Define
\[
  A_0=A_1=1,\qquad
  A_{n+1}=A_n\left(1+\frac{a_n}{n}\right),
\]
The product \(A_n\) is the first invariant of the scheme. The second invariant
is the bracket of the martingale \(M_n=S_n/A_n\), denoted by
\[
  V_n=\langle M\rangle_n.
\]
The pair \((A_n,V_n)\) is the canonical product--bracket geometry.

All convergence assumptions in the master theorem are annealed statements unless
explicitly stated otherwise. In the deterministic, or externally sampled and then
conditioned, case the same assumptions may be read quenched, path by path. This
distinction matters only for how the product--bracket geometry is verified; the
martingale classification itself is identical.

The predictable elephant random walk is the main example. It has increments
\(X_n\in\{-1,1\}\), position \(S_n=\sum_{j=1}^nX_j\), and \(S_0=0\). At time
\(n+1\), choose \(K_n\) conditionally uniform on \(\{1,\ldots,n\}\). If
\(p_n\) is \(\F_n\)-measurable and \(a_n=2p_n-1\), then conditional on
\(\F_n\) and \(K_n\),
\[
  X_{n+1}=
  \begin{cases}
  X_{K_n}, & \text{with probability }p_n,\\
  -X_{K_n}, & \text{with probability }1-p_n.
  \end{cases}
\]
Consequently
\[
  \E[X_{n+1}\mid\F_n]=a_n\frac{S_n}{n},
\]
so the ERW is a predictable linear reinforcement scheme. In this case the
bracket has the explicit decomposition
\[
  V_n=Q_n-R_n,\qquad
  Q_n=\sum_{k=1}^{n-1}\frac1{A_{k+1}^2},\qquad
  R_n=\sum_{k=1}^{n-1}\frac{a_k^2S_k^2}{k^2A_{k+1}^2}.
\]
Thus the ERW phase diagram reduces to proving that \(R_n\) is negligible
relative to the product-determined clock \(Q_n\).

\begin{definition}[Admissible product--bracket geometry]\label{def:admissible}
A predictable linear reinforcement scheme is called admissible on \([0,1]\) if
\(V_n\to\infty\)
in probability
and there exist deterministic functions \(r,q:(0,1]\to\R\), with \(q\)
nondecreasing and \(r(1)=q(1)=1\), such that for every \(\eta>0\),
\[
  \sup_{\eta\le t\le1}
  \left|\frac{A_{\lfloor nt\rfloor}}{A_n}-r(t)\right|
  \xrightarrow{\Pbb}0,
  \qquad
  \sup_{\eta\le t\le1}
  \left|\frac{V_{\lfloor nt\rfloor}}{V_n}-q(t)\right|
  \xrightarrow{\Pbb}0,
\]
\[
  \max_{k\le n}\frac{|\Delta M_k|}{\sqrt{V_n}}\xrightarrow{\Pbb}0.
\]
It is called tight at the origin if, for every \(\delta>0\),
\[
  \lim_{\eta\downarrow0}\limsup_{n\to\infty}
  \Pbb\left(
  \sup_{0\le t\le\eta}
  \frac{|S_{\lfloor nt\rfloor}|}{A_n\sqrt{V_n}}>\delta
  \right)=0.
\]
\end{definition}

\begin{remark}
The definition is deliberately formulated in terms of \(A_n\) and \(V_n\), not
in terms of a limiting coefficient \(a\). This is the point of the classification:
the coefficient path is not itself the invariant. Two paths with the same limit
may generate different products, and two paths with different local behaviour
may nevertheless lie in the same product--bracket class.
\end{remark}

\section{The Classification Theorem}\label{sec:classification}

\begin{theorem}[Master classification]\label{thm:classification}
Let \((S_n)\) be a square-integrable predictable linear reinforcement scheme
with bounded environment \((a_n)\).
\begin{enumerate}
  \item If the scheme is admissible and tight at the origin in the sense of
  Definition~\ref{def:admissible}, and if the process
  \(G(t)=r(t)B(q(t))\), \(t>0\), \(G(0)=0\), has a continuous version, then the
  linearly interpolated walk satisfies
  \[
    \left(\frac{S_{\lfloor nt\rfloor}}{A_n\sqrt{V_n}}\right)_{0\le t\le1}
    \Rightarrow
    \bigl(G(t)\bigr)_{0\le t\le1}
  \]
  in \(C[0,1]\), where \(B\) is a standard Brownian motion.
  \item If \(V_\infty=\lim_n V_n<\infty\) almost surely and
  \(\E[V_\infty]<\infty\), then
  \(M_n=S_n/A_n\) converges almost surely and in \(L^2\) to a finite random
  variable \(M_\infty\). Consequently
  \[
    \frac{S_n}{A_n}\longrightarrow M_\infty
    \qquad\text{almost surely and in }L^2.
  \]
\end{enumerate}
\end{theorem}

Thus every limit in the paper is obtained by identifying the asymptotic geometry
of \(A_n\) and \(V_n\). The next result records the rigidity built into this
classification.

\begin{theorem}[Rigidity of the product--bracket geometry]\label{thm:rigidity}
Consider two predictable linear reinforcement schemes \(e=1,2\), with canonical
quantities \((A_n^{(e)},V_n^{(e)})\). Suppose both are admissible and tight at the origin,
and
\[
  \frac{A_n^{(1)}}{A_n^{(2)}}\xrightarrow{\Pbb} c_A\in(0,\infty),
  \qquad
  \frac{V_n^{(1)}}{V_n^{(2)}}\xrightarrow{\Pbb} c_V\in(0,\infty),
\]
and that their local ratios have the same limits:
\[
  \frac{A_{\lfloor nt\rfloor}^{(e)}}{A_n^{(e)}}\xrightarrow{\Pbb} r(t),
  \qquad
  \frac{V_{\lfloor nt\rfloor}^{(e)}}{V_n^{(e)}}\xrightarrow{\Pbb} q(t),
  \qquad e=1,2.
\]
Then both environments have the same functional limit after their own canonical
normalisation. More precisely, if
\[
  G(t)=r(t)B(q(t)),\qquad t>0,\qquad G(0)=0,
\]
has a continuous version, then for \(e=1,2\),
\[
  \frac{S_{\lfloor nt\rfloor}^{(e)}}{A_n^{(e)}\sqrt{V_n^{(e)}}}
  \Rightarrow G(t)
  \qquad\text{in }C[0,1].
\]
In particular, the scaling limit is a function only of the product--bracket
geometry \((r,q)\), not of the particular predictable rule that generated it.
\end{theorem}

\begin{remark}
The theorem is not stated as a global if-and-only-if theorem. Without additional
regularity, failure of product or bracket equivalence does not by itself specify
which alternative limit appears. Section~\ref{sec:optimality} gives explicit
families showing that the invariants are indispensable and identifies the
controlled classes where converses are valid.
\end{remark}

\section{Martingale and Bracket Theory}\label{sec:proofs}

\begin{lemma}[Exact martingale normalisation]\label{lem:martingale}
The process \(M_n=S_n/A_n\) is a martingale with respect to \((\F_n)\), and
\[
  \Delta M_{n+1}
  =
  \frac{\xi_{n+1}-a_nS_n/n}{A_{n+1}}.
\]
\end{lemma}

\begin{proof}
By the defining drift condition,
\[
  \E[S_{n+1}\mid\F_n]
  =S_n+a_n\frac{S_n}{n}
  =\left(1+\frac{a_n}{n}\right)S_n.
\]
Dividing by \(A_{n+1}=A_n(1+a_n/n)\) gives
\(\E[M_{n+1}\mid\F_n]=M_n\). The increment formula follows by subtracting
\(M_n\).
\end{proof}

\begin{lemma}[ERW bracket decomposition]\label{lem:bracket-decomposition}
For the predictable elephant random walk, for all \(n\ge1\),
\[
  \langle M\rangle_n
  =
  \sum_{k=1}^{n-1}
  \frac{1-a_k^2S_k^2/k^2}{A_{k+1}^2}
  =
  Q_n-R_n.
\]
\end{lemma}

\begin{proof}
By Lemma~\ref{lem:martingale},
\[
  \E[(\Delta M_{k+1})^2\mid\F_k]
  =
  \frac{\E[(X_{k+1}-a_kS_k/k)^2\mid\F_k]}{A_{k+1}^2}.
\]
Since \(X_{k+1}^2=1\) and
\(\E[X_{k+1}\mid\F_k]=a_kS_k/k\), the conditional variance equals
\(1-a_k^2S_k^2/k^2\). Summation gives the formula.
\end{proof}

\begin{lemma}[ERW diffusive second moment bound]\label{lem:second-moment}
For the predictable elephant random walk, if \(a_n\le \bar a<1/2\) for all
\(n\), then
\[
  \sup_{n\ge1}\frac{\E[S_n^2]}{n}<\infty.
\]
\end{lemma}

\begin{proof}
Let \(W_n=\E[S_n^2]\). From the conditional drift,
\[
  \E[S_{n+1}^2\mid\F_n]
  =
  \left(1+\frac{2a_n}{n}\right)S_n^2+1
  \le
  \left(1+\frac{2\bar a}{n}\right)S_n^2+1.
\]
Iteration of this scalar recursion gives
\[
  W_n\le C n^{2\bar a}
  \left(1+\sum_{k<n}k^{-2\bar a}\right)\le C n,
\]
because \(2\bar a<1\).
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:classification}]
Assume first that the environment is admissible and tight at the origin. Fix
\(\eta>0\). On \([\eta,1]\), the martingale
\[
  N_n(t)=\frac{M_{\lfloor nt\rfloor}-M_1}{\sqrt{V_n}}
\]
has jumps bounded by
\[
  \max_{k\le n}\frac{|\Delta M_k|}{\sqrt{V_n}},
\]
which tend to zero by admissibility. Its predictable bracket is
\[
  \langle N_n\rangle(t)
  =
  \frac{V_{\lfloor nt\rfloor}}{V_n}
  \xrightarrow{\Pbb} q(t)
  \qquad\text{uniformly for }\eta\le t\le1.
\]
The martingale functional central limit theorem
\citep[Chapter 3]{hall1980martingale} yields
\[
  N_n\Rightarrow B(q(t))
  \qquad\text{on }[\eta,1].
\]
Moreover \(A_nM_1/(A_n\sqrt{V_n})=M_1/\sqrt{V_n}\to0\), so
\[
  \frac{S_{\lfloor nt\rfloor}}{A_n\sqrt{V_n}}
  =
  \frac{A_{\lfloor nt\rfloor}}{A_n}
  \frac{M_{\lfloor nt\rfloor}}{\sqrt{V_n}}
  \Rightarrow r(t)B(q(t))
  \qquad\text{on }[\eta,1]
\]
by the continuous mapping theorem. The tightness-at-the-origin condition makes
the contribution of \(t\le\eta\) uniformly negligible after \(\eta\downarrow0\).
Since the asserted limit has a continuous version with value \(0\) at the
origin, convergence holds in \(C[0,1]\) for the linearly interpolated walk.

If \(V_\infty<\infty\), then
\[
  \sup_n\E[M_n^2]
  =
  \E[M_1^2]+\sup_n\E[\langle M\rangle_n]
  \le
  \E[M_1^2]+\E[V_\infty]<\infty.
\]
The martingale convergence theorem gives \(M_n\to M_\infty\) almost surely and
in \(L^2\).
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:rigidity}]
Apply Theorem~\ref{thm:classification} to each environment. The hypotheses give
the same limiting functions \(r\) and \(q\). The constants \(c_A\) and \(c_V\)
change the canonical normalising constants but disappear because each walk is
normalised by its own \(A_n^{(e)}\sqrt{V_n^{(e)}}\). Hence both limits are the
same process \(r(t)B(q(t))\).
\end{proof}

\section{The Phase Diagram}\label{sec:regimes}

\subsection{Diffusive environments}

\begin{definition}[Uniform diffusive product regularity]\label{def:diff-reg}
An environment is uniformly diffusive with exponent \(a<1/2\) if \(a_n\le\bar
a<1/2\) and, for every \(\alpha>0\), there is a deterministic constant
\(C_\alpha<\infty\) such that, with probability tending to one,
\[
  \frac{A_n}{A_k}\le C_\alpha\left(\frac{n}{k}\right)^{a+\alpha},
  \qquad
  \frac{A_k}{A_n}\le C_\alpha\left(\frac{k}{n}\right)^{a-\alpha}
  \qquad(1\le k\le n).
\]
\end{definition}

\begin{lemma}[Product classification below criticality]\label{lem:product-conv}
For a deterministic environment, assume \(a_n\to a<1/2\), \(a_n\le\bar a<1/2\),
and \(-1+\varepsilon\le a_n\le1-\varepsilon\). Then
\[
  A_n=n^a\ell_n,
\]
where \((\ell_n)\) is slowly varying:
\[
  \frac{\ell_{\lfloor nt\rfloor}}{\ell_n}\to1
  \qquad(t>0).
\]
Consequently
\[
  Q_n\sim \frac{n}{(1-2a)A_n^2},
  \qquad
  A_n\sqrt{Q_n}\sim \sqrt{\frac{n}{1-2a}}.
\]
\end{lemma}

\begin{proof}
The identity
\[
  \log\frac{A_n}{n^a}
  =
  \sum_{k<n}\frac{a_k-a}{k}+O(1)
\]
has a summable Taylor remainder because \(a_k\) is bounded away from \(-1\).
For fixed \(t>0\), the difference between the same expression at
\(\lfloor nt\rfloor\) and at \(n\) is bounded by
\[
  \sup_{k\ge n(t\wedge1)/2}|a_k-a|
  \sum_{k\asymp n}\frac1k+o(1),
\]
which tends to zero. Thus \(A_n/n^a\) is slowly varying. Karamata's theorem for
regularly varying sequences gives
\[
  \sum_{k<n}\frac1{A_{k+1}^2}
  \sim
  \frac{n}{(1-2a)A_n^2}.
\]
\end{proof}

\begin{corollary}[Diffusive ERW classification]\label{cor:diffusive}
For the predictable elephant random walk, assume that \(A_n=n^a\ell_n\) with
\(\ell_n\) slowly varying, that the
environment is uniformly diffusive in the sense of Definition~\ref{def:diff-reg},
and that \(a_n\le\bar a<1/2\). Then
\[
  \frac{S_n}{\sqrt n}\Rightarrow
  \N\left(0,\frac1{1-2a}\right).
\]
If, in addition, the product ratios satisfy
\[
  \frac{A_{\lfloor nt\rfloor}}{A_n}\to t^a
  \]
locally uniformly on \((0,1]\), then the linearly interpolated process satisfies
\[
  \left(\frac{S_{\lfloor nt\rfloor}}{\sqrt n}\right)_{0\le t\le1}
  \Rightarrow
  \left(t^aB\left(\frac{t^{1-2a}}{1-2a}\right)\right)_{0\le t\le1}.
\]
\end{corollary}

\begin{proof}
Regular variation gives
\(Q_n\sim n\{(1-2a)A_n^2\}^{-1}\) in probability. Let \(E_n\) be the
high-probability event where the uniform product bounds hold. On \(E_n\), the
self-interaction correction satisfies
\[
  \E\left[\frac{R_n}{Q_n}\1_{E_n}\right]
  \le
  C n^{2a+2\alpha-1}
  \sum_{k<n}\frac{\E[S_k^2]}{k^{2a+2\alpha+2}}
  \le
  C n^{2a+2\alpha-1}
  \sum_{k<n}k^{-2a-2\alpha-1}
  \to0
\]
after choosing \(\alpha>0\) with \(a+\alpha<1/2\). Since \(0\le R_n/Q_n\le1\), this
proves \(R_n/Q_n\to0\) in probability after adding the probability of the
complementary event. The same estimate with upper limit \(\lfloor nt\rfloor\)
gives
\[
  \sup_{\eta\le t\le1}\frac{R_{\lfloor nt\rfloor}}{Q_n}
  \xrightarrow{\Pbb}0
  \qquad(\eta>0).
\]
By Lemma~\ref{lem:bracket-decomposition}, \(V_n=Q_n-R_n\), so \(V_n/Q_n\to1\)
in probability and the same local ratios hold for \(V\).
Moreover,
\[
  \frac{Q_{\lfloor nt\rfloor}}{Q_n}\to t^{1-2a}
\]
locally uniformly on \((0,1]\), and the assumed product-ratio convergence gives
\(r(t)=t^a\). The jump condition follows from
\[
  \max_{k\le n}\frac{|\Delta M_k|}{\sqrt{V_n}}
  \le
  2\max_{k\le n}\frac1{A_{k+1}\sqrt{V_n}}\xrightarrow{\Pbb}0.
\]
Finally, for \(m=\lfloor n\eta\rfloor\), Lemma~\ref{lem:second-moment} and
Doob's inequality applied to the martingale \(M\) give
\[
  \E\left[
  \sup_{0\le t\le\eta}
  \left(\frac{S_{\lfloor nt\rfloor}}{A_n\sqrt{Q_n}}\right)^2
  \right]
  \le C\eta.
\]
Since \(V_n/Q_n\to1\) in probability, this gives tightness at the origin with
the canonical denominator \(A_n\sqrt{V_n}\).
Thus the environment is admissible and tight at the origin. Theorem
\ref{thm:classification} yields the functional limit after canonical
normalisation; multiplying by
\(A_n\sqrt{V_n}/\sqrt n\to(1-2a)^{-1/2}\) in probability gives the stated CLT
and FCLT.
\end{proof}

\subsection{Critical environments}

\begin{corollary}[Critical ERW classification]\label{cor:critical}
For the predictable elephant random walk, assume that there is a deterministic
constant \(D_c<\infty\) such that
\[
  \sum_{n=1}^{\infty}\frac{|a_n-1/2|}{n}\le D_c,
  \qquad a_n\le\frac12,
\]
and \(-1+\varepsilon\le a_n\le1-\varepsilon\). Then
\[
  A_n=L n^{1/2}(1+o(1))
  \quad\text{for some }L\in(0,\infty),
  \qquad
  Q_n\sim L^{-2}\log n.
\]
Consequently
\[
  \frac{S_n}{\sqrt{n\log n}}\Rightarrow \N(0,1),
\]
and the linearly interpolated process satisfies
\[
  \left(\frac{S_{\lfloor nt\rfloor}}{\sqrt{n\log n}}\right)_{0\le t\le1}
  \Rightarrow
  \left(\sqrt t\,N\right)_{0\le t\le1}
  \quad\text{in }C[0,1],
\]
where \(N\sim\N(0,1)\).
\end{corollary}

\begin{proof}
The product estimate is the logarithmic analogue of
Lemma~\ref{lem:product-conv}; the summability assumption makes
\(A_n/n^{1/2}\) converge to a finite positive limit. Hence
\[
  Q_n=\sum_{k<n}A_{k+1}^{-2}\sim L^{-2}\log n.
\]
The second moment recursion gives \(E[S_n^2]\le Cn\log n\). Therefore
\[
  \E\left[\frac{R_n}{Q_n}\right]
  \le
  \frac{C}{\log n}\sum_{k<n}\frac{\log k}{k^2}\to0.
\]
For every \(\eta>0\), uniformly for \(t\in[\eta,1]\),
\[
  \frac{A_{\lfloor nt\rfloor}}{A_n}\to\sqrt t,
  \qquad
  \frac{Q_{\lfloor nt\rfloor}}{Q_n}\to1.
\]
By Lemma~\ref{lem:bracket-decomposition}, \(V_n/Q_n\to1\) in probability and the
same local bracket ratios hold for \(V\).
The jump condition follows from
\[
  \max_{k\le n}\frac{|\Delta M_k|}{\sqrt{V_n}}
  \le \frac{C}{\sqrt{\log n}}\to0,
\]
and the preceding estimate on \(R_n/Q_n\) also holds uniformly on
\([\eta,1]\). The origin condition is
\[
  \E\left[\sup_{0\le t\le\eta}
  \left(\frac{S_{\lfloor nt\rfloor}}{\sqrt{n\log n}}\right)^2\right]
  \le C\eta\frac{\log(n\eta+2)}{\log n}.
\]
Since \(A_n\sqrt{V_n}\sim\sqrt{n\log n}\) in probability, this is exactly the tightness at the
origin required by Definition~\ref{def:admissible}. The limiting process is
\(G(t)=\sqrt t\,B(1)\), with \(G(0)=0\), which is continuous. Theorem
\ref{thm:classification} completes the proof.
\end{proof}

\subsection{Superdiffusive environments}

\begin{corollary}[Superdiffusive ERW classification]\label{cor:super}
For the predictable elephant random walk, if \(a_n\to a>1/2\) and there is a
deterministic constant \(D_s<\infty\) such
that
\[
  \sum_{n=1}^\infty \frac{|a_n-a|}{n}\le D_s,
\]
then \(A_n=L n^a(1+o(1))\) for some \(L\in(0,\infty)\), \(V_\infty<\infty\),
and
\[
  \frac{S_n}{n^a}\longrightarrow L M_\infty
  \qquad\text{almost surely and in }L^2.
\]
\end{corollary}

\begin{proof}
The product estimate gives \(A_n=L n^a(1+o(1))\). The deterministic bound
\(\sum |a_n-a|/n\le D_s\), together with the bounded Taylor remainders in
\(\log(1+a_n/n)\), also gives a deterministic lower bound
\[
  A_n\ge c n^a
  \qquad(n\ge1)
\]
for some \(c>0\). Hence
\[
  Q_\infty\le c^{-2}\sum_{n\ge1}n^{-2a}<\infty
\]
deterministically. Since \(V_n\le Q_n\), also \(V_\infty<\infty\) and
\(\E[V_\infty]<\infty\). The finite-bracket part of
Theorem~\ref{thm:classification} applies.
\end{proof}

\section{Why the Classification Is Optimal}\label{sec:optimality}

\begin{proposition}[Regularly varying ERW clock diagram]\label{prop:rv-phase}
For the predictable elephant random walk, suppose that the product of a bounded
environment satisfies
\[
  A_n\sim C n^\alpha L(n),
\]
where \(C\in(0,\infty)\), \(L\) is slowly varying, and the self-interaction
correction is negligible whenever \(Q_n\to\infty\). Then the product clock
\(Q_n=\sum_{k<n}A_{k+1}^{-2}\), and hence the ERW bracket whenever
\(R_n=o(Q_n)\), has the following three regimes.
\begin{enumerate}
  \item If \(\alpha<1/2\), then
  \[
    Q_n\sim \frac{n}{(1-2\alpha)A_n^2},
    \qquad
    A_n\sqrt{Q_n}\sim\sqrt{\frac{n}{1-2\alpha}},
  \]
  so the canonical Gaussian scale is diffusive.
  \item If \(\alpha=1/2\), then
  \[
    Q_n\sim C^{-2}\sum_{k<n}\frac1{kL(k)^2}.
  \]
  The critical normalisation is therefore determined by the slowly varying
  integral, not by the exponent alone.
  \item If \(\alpha>1/2\), then \(Q_\infty<\infty\), and the walk is in the
  finite-bracket martingale regime.
\end{enumerate}
\end{proposition}

\begin{proof}
This is Karamata's theorem applied to
\(A_{k+1}^{-2}\sim C^{-2}k^{-2\alpha}L(k)^{-2}\). The cases
\(\alpha<1/2\), \(\alpha=1/2\), and \(\alpha>1/2\) correspond respectively to a
divergent power sum, a borderline slowly varying sum, and a convergent sum. The
classification theorem then translates these bracket alternatives into the
corresponding limit regimes.
\end{proof}

\begin{proposition}[Coefficient convergence does not classify products]\label{prop:slow}
Fix \(a<1/2\) and \(c\in\R\). The deterministic environment
\[
  a_n=a+\frac{c}{\log(n+2)}
\]
can be chosen bounded away from \(-1\) and \(1\), satisfies \(a_n\to a\), and
\[
  A_n\sim C_c n^a(\log n)^c
  \qquad\text{for some }C_c\in(0,\infty).
\]
Thus the limiting coefficient \(a\) does not determine the martingale
normalisation.
\end{proposition}

\begin{proof}
The Taylor expansion of the logarithm gives
\[
  \log A_n
  =
  \sum_{k<n}\frac{a}{k}
  +c\sum_{k<n}\frac1{k\log(k+2)}
  +O(1)
  =
  a\log n+c\log\log n+O(1).
\]
Exponentiating proves the claim.
\end{proof}

\begin{proposition}[Different invariants give different canonical laws]\label{prop:different}
Let \(a<1/2\). For the constant environment \(a_n\equiv a\) and the slow
environment \(a_n=a+c/\log(n+2)\), \(c\ne0\), the terminal diffusive CLT has the
same variance \((1-2a)^{-1}\), but the canonical martingale normalisations differ
by the factor \((\log n)^c\). Hence no classification based only on the limiting
coefficient can be pathwise complete.
\end{proposition}

\begin{proof}
Both environments satisfy Lemma~\ref{lem:product-conv}, so
Corollary~\ref{cor:diffusive} gives the same terminal CLT. Proposition
\ref{prop:slow} shows that their products are not asymptotically equivalent
unless \(c=0\). Since \(S_n=A_nM_n\), the martingale-scale description retains
this distinction even when the terminal \(\sqrt n\)-CLT does not.
\end{proof}

\begin{proposition}[Logarithmic critical trichotomy]\label{prop:critical-window}
Let
\[
  a_n=\frac12+\frac{c}{\log(n+2)}
\]
for all large \(n\). Then
\[
  A_n\sim C n^{1/2}(\log n)^c.
\]
Consequently the bracket scale is
\[
  Q_n\asymp
  \begin{cases}
    (\log n)^{1-2c}, & c<1/2,\\
    \log\log n, & c=1/2,\\
    1, & c>1/2.
  \end{cases}
\]
Consequently the canonical normalisation \(A_n\sqrt{Q_n}\) has the following
orders:
\[
  A_n\sqrt{Q_n}\asymp
  \begin{cases}
    \sqrt{n\log n}, & c<1/2,\\
    \sqrt{n\log n\log\log n}, & c=1/2,\\
    \sqrt n(\log n)^c, & c>1/2.
  \end{cases}
\]
Thus the boundary is not determined by the pointwise limit \(a_n\to1/2\), but by
the product--bracket geometry; the same limiting coefficient produces diffusive
logarithmic scaling, iterated-logarithmic critical scaling, or finite-bracket
supercritical behaviour according to the value of \(c\).
\end{proposition}

\begin{proof}
The product asymptotic is the calculation of Proposition~\ref{prop:slow} with
\(a=1/2\). The displayed alternatives follow from the integral test applied to
\[
  Q_n\sim C^{-2}\sum_{k<n}\frac1{k(\log k)^{2c}}.
\]
Multiplying these estimates by \(A_n^2\sim C^2n(\log n)^{2c}\) gives the
orders of \(A_n\sqrt{Q_n}\).
\end{proof}

\begin{remark}
These examples are deliberately deterministic. They show that the obstruction is
not caused by random learning noise or measurability subtleties. It is already
present in the product geometry of the simplest predictable environments.
\end{remark}

\section{Examples of Predictable Environments}\label{sec:examples}

Learning rules are not part of the main theory; they are sources of predictable
environments. The classification theorem applies once the product and bracket
geometry has been identified.

\begin{proposition}[Lipschitz policy maps]\label{prop:lipschitz}
Let \(\theta_n\) be \(\F_n\)-measurable, let
\(\ell:\R^d\to(\varepsilon,1-\varepsilon)\) be Lipschitz, and put
\[
  p_n=\ell(\theta_n),\qquad a_n=2p_n-1.
\]
If \(\theta_n\to\theta_\star\) almost surely, \(a=2\ell(\theta_\star)-1<1/2\),
\(a_n\le\bar a<1/2\), and the induced product is uniformly diffusive in the
sense of Definition~\ref{def:diff-reg}, then the diffusive classification
applies. If there is a deterministic constant \(D_\theta<\infty\) such that
\[
  \sum_{n=1}^\infty \frac{\|\theta_n-\theta_\star\|}{n}\le D_\theta,
\]
then the corresponding stable diffusive, critical or superdiffusive product
asymptotics follow according to the value of \(a\).
\end{proposition}

\begin{proof}
Lipschitz continuity gives
\[
  |a_n-a|\le 2\,\operatorname{Lip}(\ell)\|\theta_n-\theta_\star\|.
\]
The claims are direct applications of the phase diagram.
\end{proof}

\begin{remark}[On regret and stochastic approximation]
Algorithms such as projected stochastic approximation, confidence-bound
procedures, exponential weights or posterior sampling enter the present theory
only after they have been proved to generate a predictable coefficient path with
one of the product--bracket geometries above. The paper does not use regret as a
black box and does not claim a universal theorem for these algorithms.
\end{remark}

\begin{remark}[Numerical diagnostics]
The accompanying scripts reproduce finite-sample checks for the diffusive,
critical, superdiffusive and slowly varying regimes. These computations are not
used as evidence for the theorems; they are retained only to make the examples
and normalisations reproducible.
\end{remark}

\section{Discussion and Open Problems}\label{sec:open}

The classification suggests several problems beyond the scalar predictable
environment treated here.

\paragraph{Multidimensional predictable environments.}
For walks on \(\mathbb Z^d\) with increments taking values among the coordinate
directions and their opposites, the same scalar product \(A_n\) normalises the
conditional drift whenever
\(\E[X_{n+1}\mid\F_n]=a_nS_n/n\). The remaining issue is the limiting
directional covariance of the martingale increments. A complete vector
classification should combine the scalar product--bracket geometry with a
matrix-valued covariance invariant.

\paragraph{Continuous-time analogues.}
The product \(A_n\) is the discrete solution of a logarithmic-time linear
equation. Continuous-time memory processes should lead to multiplicative
functionals of predictable drift environments and bracket clocks of the same
kind.

\paragraph{Random memory kernels.}
The present paper keeps the recalled time uniform and puts all randomness in the
repeat--reverse environment. If the recall kernel itself becomes predictable,
the conditional drift is no longer \(a_nS_n/n\). Identifying the right invariants
for that problem would extend the classification beyond scalar products.

\paragraph{Quenched and annealed universality.}
The results here are pathwise in the realised predictable environment. A more
ambitious theory would distinguish quenched limits conditional on an external
environment from annealed limits when the environment has its own dynamics.
Feedback environments, where \(a_n\) is generated from the walk itself, are the
main obstruction to a direct random-media analogy.

\paragraph{Reinforced and self-interacting models.}
The product--bracket mechanism may also appear in reinforced walks or adaptive
urns after linearisation around a stable direction. The open question is whether
there is a broader martingale classification principle behind these models, or
whether the complete-memory ERW is a special solvable case.

\bibliographystyle{apalike}
\bibliography{references}

\end{document}
